DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2022-05-23 manuscript/manuscript.Rmd
BCO comments and mild edits

  Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA.

abstract: |
  Date estimates for times of evolutionary divergences are key data for research in the natural sciences. These estimates also provide valuable information for for education, science communication and policy decisions.
  Although achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
  When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
  Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
  Here we present DateLife, a service implemented as an R package and an Rshiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
  The main DateLife workflow constructs a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from TreeBASE database as well.
  We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
  DateLife will be useful to increase awereness of the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypotheses on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.

@@

DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree. If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004]. The input tree is not required to have branch lengths, and its topology is used in the summary steps described below.

DateLife accepts scientific names that can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or a binomial specific. Subspecies and variants are ignored. If an input taxon name belongs to an inclusive taxonomic group the algorithm has two alternative behaviors defined by the "get species from taxon" flag. If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (according to a taxonomy) and adds them to the input string. If the flag is inactive, DateLife excludes the taxon names above the species level from the input.

DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database. TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard [@Boyle2013]. DateLife implements TNRS using OpenTree's unified taxonomy as standard [@opentreeAPIs; @rees2017automated], storing OpenTree's Taxonomy identification numbers for further processing.
<!-- We are using OpenTree's TNRS, Boyle described TNRS first, right?  -->

The processed input taxon names are saved as an R object of a newly defined class \texttt{datelifeQuery} that is used in the following steps. This object contains the standardized names, the corresponding taxonomic id numbers, and the topology of the input tree if any was provided.

@@

DateLife's chronogram database latest version consist of `r length(opentree_chronograms$trees)` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata [@mctavish2015phylesystem], which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the `datelife` R package.
Published chronograms can be added to Phylesystem by any user, at any time, and are immediately publicly available (https://tree.opentreeoflife.org/curator). This facilitates an immediate update of DateLife's chronogram database to include new chronogram data on a following search.

A DateLife search is implemented by matching processed taxon names provided by the user, to tip labels in the chronogram database. Chronograms with at least two matching taxon names on their tip labels are identified and pruned down to preserve only the matched taxa.
These matching pruned chronograms are referred to as source chronograms.
Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a "phylo" object or newick string.
Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.

@@

_**Dating the topology.--**_ 
Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind.
Optionally, to estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for the input taxa.
These markers are aligned with MUSCLE by default [@edgar2004muscle], or MAFFT [@katoh2009multiple]. 
This alignment can be used to estimate branch lengths on input topologies that lack branch lengths.
Currently, branch length reconstruction in DateLife is performed using parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed [@schliep2011phangorn].
While relative branch length information provides additional data for nodes without secondary date calibrations, topologies without branch lengths can also be dated.

Once a topology is chosen, DateLife applies the congruification method [@Eastman2013] to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as `datelifeResult`. By definition, the matrices store total distance (time from tip to tip), hence, node ages correspond to half the values stored in the patristic distance matrices.
A table of congruified node ages that can be used as secondary calibrations is stored as a `congruifiedCalibrations` object.

@@

By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology. BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
This minimizes age variance in the resulting chronogram [@Webb2008].
BLADJ does not incorporate branch lengths even when they are present.
When there is conflict in ages between nodes with calibration data, the algorithm ignores node ages that are older than the age of a parent node. 
BLADJ requires a root age estimate. If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature. If none is provided, DateLife assigns an arbitrary age to the root as 10% older than the oldest age available within the group.

Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the secondary calibrations.
PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees.
treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
[@Smith2012].
The MrBayes  [@Huelsenbeck2001; @Ronquist2003] approach in datelife uses the secondary calibrations as priors on node ages.

@@

`datelife`'s code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
Query processing time increases roughly linearly with number of input taxon names, and
increases considerably if Taxonomic Name Resolution Service (TNRS) is activated.
<!-- Defining tnrs here again, bc we only talk about it two times, so it easy to forget it -->
Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig. \ref{fig:runtime_main}).
<!--Speed of other datelife workflows? NO, bc it reflects speed from other software mainly. But I will put a sentence about it...
Summarizing `datelife` results processing times: shows sdm speed, clustering methods speed;
generating new trees: shows bladj speed;
Adding dates processing time: reflects more mrbayes speed than anything;
get_bold_otol_tree running time: reflects bold downloading data speed.-->
<!--Speed with different number of chronograms in database? NO-->
<!--Speed of web interface and of r package in computers with different capacities? NOOOO-->
`datelife`'s code performance was evaluated with a set of unit tests designed and
implemented with the R package testthat [@RCoreTeam2018] that were run both locally
with the devtools package [@RCoreTeam2018], and on a public server --via
GitHub, using the continuous integration tool Travis CI (<https://travis-ci.org>). At
present, unit tests cover more than 40% of `datelife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->

@@

We illustrate the DateLife algorithm using a group within the Passeriform birds encompassing the family of true finches, Fringillidae as case study. The first example analyses 6 bird species and shows all steps of the algorithm. The second example is an application analysing all 289 species in the family Fringillidae for which we have date informtaion.
<!--We used `r cite_r("r-references.bib")` for all our analyses.-->

@@

Processing input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the default reference taxonomy [Open Tree of Life Taxonomy v3.3, June 1, 2021]. For a detailed discussion on the state of the synonym refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza]. 
DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies. Three studies matched five input names [@barker2015new; @Hedges2015; @Jetz2012], one study matched four input names [@hooper2017chromosomal] and two studies matched two input names [@barker2013going; @burns2014phylogenetics]. No studies matched all input names. Together, matching chronograms have 28 unique age data points. All nodes have age data.
DateLife used OpenTree's synthetic tree topology for these taxa and mapped age data to nodes in the tree.
As expected, more inclusive nodes (e.g., node "n1") have more age data than less inclusive nodes (e.g., node "n5").
The processing step allowed discovering five data points for node "n4" that would not have had any data otherwise. \textcolor{red}{I don't understand what this sentence means}
Age summary statistics per node were calculated and used as secondary calibrations to date the tree topology using the BLADJ algorithm.
Age data for node "n2" was excluded as final calibration because it is older than age data of a more inclusive node (n1).

@@

To estimate ages for the taxa in the family Fringillidae, we ran a `datelife` query using the "get species from taxon" flag. 
This will get all recognised species names within the named taxon. 
The Fringillidae has species, according to the Open Tree of Life taxonomy.
??? We have date estimates for `r length(dquery$cleaned_names)` of those species.

??? There is an option to include all, right?
According to the OpenTree of life synthetic phylogeny, Fringillidae is not a monophyletic family, so this taxon-constrained approach returns a topology and ages for the species in that named group.

The next step is to run a `datelife` search on this curated set of species taxon names that will find all chronograms that contain at least two species. The algorithm proceeds to prune the trees to keep matching species names on tips only, and transform the pruned trees to pairwise distance matrices.
There are `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
The final step is to summarize the available information using two alternative types of summary chronograms, median and SDM. As explained in the "Description" section, data from source chronograms is first summarised into a single distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig. \ref{fig:fringillidages}). Median sxummary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
<!-- ????????????????? Say some things about the results!
ANSWER:
(Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
<!-- and hierarchical clustering method which appears to push ages back -->

@@

In 5 studies, more inclusive nodes have older original ages, and less inclusive nodes have younger original ages than their cross-validated age estimates. 
\textcolor{red}{more inclusive nodes have to be older! :P I think there is something differnet you are trying to say here - maybe "Datelife tended to underestimate ages for more inclusive nodes relative to the orginal estimate, and overestimate ages for nodes closer to the tips"}
Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. \ref{fig:cvXYdiffs}).

@@

`datelife` makes state-of-the-art information on time of lineage divergence easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter. It is an open service that does not require any expert biological knowledge from users --besides the names of the organisms they want to work with, for any of its functionality.

At the time of writing of this manuscript (`r format(Sys.time(), "%b %d, %Y")`), `datelife`'s database has `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OpenTree's database, the Phylesystem [@mctavish2015phylesystem]. A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time. As chronograms are added to Phylesystem, they are incorporated into an updated `datelife`'s database that is assigned a new version number, followed by a package release on CRAN. `datelife`'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
Users can also upload new chronograms to OpenTree themselves, and trigger an update of their local `datelife` database to incorporate the new chronograms, to have them immediately available for analysis.

A total of `r length(unique(unname(unlist(lapply(opentree_chronograms$trees, "[[", "tip.label")))))` unique terminal taxa are represented in `datelife`'s database.
Incorporation of more chronograms into `datelife`'s database will continue to improve its services. One option to increase chronogram number in the database is the Dryad data repository.  Methods to automatically mine chronograms from Dryad could be designed and implemented. However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a curator.

The largest, and taxonomically broadest, summary chronogram currently available from OpenTree
was constructed using age data from 2,274 published chronograms [@Hedges2015]. 
Unfortunately, the source chronograms used as input data for this tree are not available in computer readable format for reuse or reanalysis.
Access to the input chronograms used to generate the Hedges et al. summary tree would improve measures of uncertainty in DateLife, but they are currently available only as image files (timetree.org).
<!--IN VIOLATION OF NSF RULES! but I don't suppose we should say that.-->
We would like to emphasize on the importance of sharing chronogram data,  including systematically curated metadata,  into open repositories, such as OpenTree's Phylesystem [@mctavish2015phylesystem] for the benefit of the scientific community as a whole.

By default, `datelife` currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
Strictly speaking, the best chronogram should reflect the real time of lineage divergence accurately and precisely.
To our knowledge, there are no good measures to determine independently if a chronogram is better than another.
Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used [@magallon2010using, @magallon2015metacalibrated].
Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms.
Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.

@@

Conflict among alternative date estimates is common in the literature
(see, for example, the ongoing debate about crown group age of angiosperms [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time]. 
Using a different chronogram may return different results [@title2016macrophylogenies] particularly for studies of species diversification, or focussed on the timing of evolutionary events.
Stitching together multiple chronograms creates a larger tree that uses information from across studies, but may amplify uncertainty and errors.

Summarizing chronograms might also imply summarizing fundamentally distinct evolutionary hypotheses.
For example, two different researchers working on the same clade both carefully select and argument their choices of fossil calibrations.
Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence [@sauquet2013practical]. Trying to summarize the resulting chronograms into a single one using simple summary statistics can erase many types of relevant information from the source chronograms. Accordingly, the prevailing view is that we should favor time of lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history [@antonelli2017supersmart]. <!-- Will this be implemented in future `datelife` versions? Probably not-->
However, the exercise of summarizing different chronograms has the potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis. Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram [@sauquet2021age]. <!-- Example -->
\textcolor{red}{??? it's not clear to me what this last part means! What might you be losing?}

>>>>

  Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 446 Hesler Biology Building, Knoxville, TN 37996, USA.

abstract: |
  Date estimates for times of evolutionary divergences are key data for research in the natural sciences. These estimates also provide valuable information for for education, science communication, and policy decisions.
  Although achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram) is a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
  When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
  Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
  Here we present DateLife, a service implemented as an R package and an R Shiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
  The main DateLife workflow constructs a chronogram for any given combination of taxon names by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from the TreeBASE database as well.
  We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
  DateLife will be useful to increase awereness of the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypotheses on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.

@@

DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string or as tip labels on a tree. If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004]. The input tree is not required to have branch lengths, and its topology is used in the summary steps described below.

DateLife accepts scientific names that can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or a binomial species name. Subspecies and variants are ignored. If an input taxon name belongs to an inclusive taxonomic group the algorithm has two alternative behaviors defined by the "get species from taxon" flag. If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (according to a taxonomy) and adds them to the input string. If the flag is inactive, DateLife excludes the taxon names above the species level from the input.

DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database. TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard [@Boyle2013]. DateLife implements TNRS using OpenTree's unified taxonomy as standard [@opentreeAPIs; @rees2017automated], storing OpenTree's Taxonomy identification numbers for further processing.
<!-- We are using OpenTree's TNRS [BCO: yes], Boyle described TNRS first, right?  [BCO: earliest I can find, but feels like there should be much older work]-->

The processed input taxon names are saved as an R object of a newly defined class \texttt{datelifeQuery} that is used in the following steps. This object contains the standardized names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.

@@

DateLife's chronogram database latest version consist of `r length(opentree_chronograms$trees)` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata [@mctavish2015phylesystem], which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the `datelife` R package.
Published chronograms can be added to Phylesystem by any user, at any time, and are immediately publicly available (https://tree.opentreeoflife.org/curator). A user may then update their local cache of DateLife's chronogram database to include new chronogram data on a following search.

A DateLife search is implemented by matching processed taxon names provided by the user to tip labels in the chronogram database. Chronograms with at least two matching taxon names on their tip labels are identified and pruned down to preserve only the matched taxa.
These matching pruned chronograms are referred to as source chronograms.
Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a "phylo" object or newick string.
Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.

@@

_**Dating the topology.--**_ 
Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind.
Optionally, to estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for the input taxa.
These markers are aligned with MUSCLE [@edgar2004muscle] (by default) or MAFFT [@katoh2009multiple]. 
This alignment can be used to estimate branch lengths on input topologies that lack branch lengths.
Currently, branch length reconstruction in DateLife is performed using parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed [@schliep2011phangorn].
While relative branch length information provides additional data for nodes without secondary date calibrations, topologies without branch lengths can also be dated.

Once a topology is chosen, DateLife applies the congruification method [@Eastman2013] to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as `datelifeResult`. By definition, the matrices store total distance (time from tip to tip), hence, node ages correspond to half the values stored in the patristic distance matrices.
This assumes that the terminal taxa are coeval and occur at the present.
A table of congruified node ages that can be used as secondary calibrations is stored as a `congruifiedCalibrations` object.

@@

By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology. BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
This minimizes age variance in the resulting chronogram [@Webb2008].
BLADJ does not incorporate input phylogeny branch lengths even when they are present.
When there is conflict in ages between nodes with calibration data, the algorithm ignores node ages that are older than the age of a parent node. 
BLADJ requires a root age estimate. If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature. If none is provided, DateLife assigns an arbitrary age to the root as 10% older than the oldest age available within the group.

Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the secondary calibrations.
PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees.
treePL [@Smith2012], is a semi-parametric, rate-smoothing, penalized likelihood dating method
[@Sanderson2002].
The MrBayes  [@Huelsenbeck2001; @Ronquist2003] approach in datelife uses the secondary calibrations as priors on node ages.

@@

`datelife`'s code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
Query processing time increases roughly linearly with number of input taxon names, and
increases considerably if Taxonomic Name Resolution Service (TNRS) is activated.
<!-- Defining tnrs here again, bc we only talk about it two times, so it easy to forget it -->
Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig. \ref{fig:runtime_main}).
<!--Speed of other datelife workflows? NO, bc it reflects speed from other software mainly. But I will put a sentence about it...
Summarizing `datelife` results processing times: shows sdm speed, clustering methods speed;
generating new trees: shows bladj speed;
Adding dates processing time: reflects more mrbayes speed than anything;
get_bold_otol_tree running time: reflects bold downloading data speed.-->
<!--Speed with different number of chronograms in database? NO-->
<!--Speed of web interface and of r package in computers with different capacities? NOOOO-->
`datelife`'s code performance was evaluated with a set of unit tests designed and
implemented with the R package testthat [@RCoreTeam2018] that were run both locally
with the devtools package [@RCoreTeam2018], and on a public server --via
GitHub, using the continuous integration tool Travis CI (<https://travis-ci.org>). At
present, unit tests cover more than 40% of `datelife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->
Unit testing helps identify potential issues as code is updated or, more critically, as services code relies upon may change.

@@

We illustrate the DateLife algorithm using a group within the Passeriform birds encompassing the family of true finches, Fringillidae, as case study. The first example analyses 6 bird species and shows all steps of the algorithm. The second example is an application analysing all 289 species in the family Fringillidae for which we have date informtaion.
<!--We used `r cite_r("r-references.bib")` for all our analyses.-->

@@

Processing input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the default reference taxonomy [Open Tree of Life Taxonomy v3.3, June 1, 2021]. For a detailed discussion on the state of the synonym refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza]. 
DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies. Three studies matched five input names [@barker2015new; @Hedges2015; @Jetz2012], one study matched four input names [@hooper2017chromosomal] and two studies matched two input names [@barker2013going; @burns2014phylogenetics]. No studies matched all input names. Together, matching chronograms have 28 unique age data points. All nodes have age data.
DateLife used OpenTree's synthetic tree topology for these taxa and mapped age data to nodes in the tree.
As expected, more inclusive nodes (e.g., node "n1") have more age data than less inclusive nodes (e.g., node "n5").
The name resolution step allowed discovering five data points for node "n4" that would not have had any data otherwise due to name mismatch. 
Age summary statistics per node were calculated and used as secondary calibrations to date the tree topology using the BLADJ algorithm.
Age data for node "n2" was excluded as final calibration because it is older than age data of a more inclusive node (n1).

@@

To estimate ages for the taxa in the family Fringillidae, we ran a `datelife` query using the "get species from taxon" flag. 
This will get all recognised species names within the named taxon. 
The Fringillidae has species, according to the Open Tree of Life taxonomy. <!--BCO: Feel like we need a number here, though it's good that the clade hasn't gone extinct ;-) -->
??? We have date estimates for `r length(dquery$cleaned_names)` of those species.

??? There is an option to include all, right?
According to the OpenTree of life synthetic phylogeny, Fringillidae is not monophyletic, so this taxon-constrained approach returns a topology and ages for the species in that named group. <!--BCO: it's ambiguous: it gets a topology and age for the larger clade that includes all Fringillidae species? Or for the paraphyletic Fringillidae group?-->

The next step is to run a `datelife` search on this curated set of species taxon names that will find all chronograms that contain at least two species. The algorithm proceeds to prune the trees to keep matching species names on tips only, and transform the pruned trees to pairwise distance matrices.
There are `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
The final step is to summarize the available information using two alternative types of summary chronograms, median and SDM. As explained in the "Description" section, data from source chronograms is first summarised into a single patristic distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig. \ref{fig:fringillidages}). Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
<!-- ????????????????? Say some things about the results!
ANSWER:
(Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
<!-- and hierarchical clustering method which appears to push ages back -->

@@

For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka 'closer to the root') relative to the orginal estimate, and overestimate ages for nodes closer to the tips. 
Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. \ref{fig:cvXYdiffs}).

@@

`datelife` makes state-of-the-art information on time of lineage divergence easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter. It is an open service that does not require any expert biological knowledge from users, besides the names of the organisms or clade they want to work with, for any of its functionality.

At the time of writing of this manuscript (`r format(Sys.time(), "%b %d, %Y")`), `datelife`'s database has `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OpenTree's database, the Phylesystem [@mctavish2015phylesystem]. A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time. As chronograms are added to Phylesystem, they are incorporated into an updated `datelife`'s database that is assigned a new version number, followed by a package release on CRAN. `datelife`'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months. <!--BCO: Nicely done-->
Users can also upload new chronograms to OpenTree themselves, and trigger an update of their local `datelife` database to incorporate the new chronograms, to have them immediately available for analysis. <!--BCO: What is immediately? Also, it's not just uploading -- they have to include metadata on what brlen mean, maybe? indicate the ingroup/outgroup, do TNRS within OToL, right? Maybe this has improved. I remember being frustrated trying to load in giant trees for datelife use b/c I had to click on the ingroup node-->
<!--BCO; this duplicates an earlier section and feels more methods-y. Maybe move up and merge?-->

A total of `r length(unique(unname(unlist(lapply(opentree_chronograms$trees, "[[", "tip.label")))))` unique terminal taxa are represented in `datelife`'s database.
Incorporation of more chronograms into `datelife`'s database will continue to improve its services. One option to increase the of chronograms in the DateLife database is the Dryad data repository, which now holds some chronograms.  Methods to automatically mine chronograms from Dryad could be designed and implemented. However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a curator.

The largest, and taxonomically broadest, summary chronogram currently available from OpenTree
was constructed using age data from 2,274 published chronograms [@Hedges2015].
<!--BCO: I just checked: TIMETREEE NOW ALLOWS SOURCE TREE DOWNLOADS! This is great news. They have a copyright (not legal in US on trees, IMHO) notice on individual trees, but they do say "To download all individual study timetrees, please send an email request to info@timetree.org." -- I think this could be great to do-->

We would like to emphasize on the importance of sharing chronogram data,  including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem [@mctavish2015phylesystem] for the benefit of the scientific community as a whole.

By default, `datelife` currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
Strictly speaking, the best chronogram should reflect the real time of lineage divergence accurately and precisely.
To our knowledge, there are no good measures to determine independently if a chronogram is better than another.
Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used [@magallon2010using, @magallon2015metacalibrated].
Several characteristics of the data used for dating analyses, as well as from the output chronogram itself, could be used to score quality of source chronograms.
Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
Its open database of trees allows other researchers to do such analyses themselves reproducibly and without needing permission, though of course they should follow proper citation practices, especially for the source trees.

@@

Conflict among alternative date estimates is common in the literature
(see, for example, the robust ongoing debate about crown group age of angiosperms [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time]. 
Using a different chronogram may return different results [@title2016macrophylogenies] particularly for studies of species diversification, or focussed on the timing of evolutionary events.
Stitching together multiple chronograms creates a larger tree that uses information from across studies, but may amplify uncertainty and errors.

Summarizing chronograms might also imply summarizing fundamentally distinct evolutionary hypotheses.
For example, two different researchers working on the same clade could both carefully select and argue for their choices of fossil calibrations.
Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence [@sauquet2013practical]. Trying to summarize the resulting chronograms into a single one using simple summary statistics can erase many types of relevant information from the source chronograms. Accordingly, the prevailing view is that we should favor lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history, as for example done by [@antonelli2017supersmart]. <!-- Will this be implemented in future `datelife` versions? Probably not. -->
There have been attempts to create fossil calibration databases <!--REF TO ADD: Ksepka, D.T., J.F. Parham, J.F. Allman, M.J. Benton, M.T. Carrano, K.A. Cranston, P.C. J. Donoghue, J.J. Head, E.J. Hermsen, R.B. Irmis, W.G. Joyce, M. Kohli, K.S. Lamm, D. Leehr, J.S.L. Patané, P.D. Polly, M.J. Phillips, N.A. Smith, N.D. Smith, M. van Tuinen, J.L. Ware, R.C.M. Warnock. 2015.The Fossil Calibration Database, A New Resource for Divergence Dating. Systematic Biology: doi:10.1093/sysbio/syv025.-->, though these still have room to grow. <!--BCO: Polite way to say 'no new additions in four years?'-->
However, the exercise of summarizing different chronograms has the potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypotheses. Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially losing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram [@sauquet2021age]. <!-- Example -->
\textcolor{red}{??? it's not clear to me what this last part means! What might you be losing?}
